A Framework for Reproducible Data Science using Nix

### 2025-10-20 paper.qmd
figures$
appendix

The `rxp_populate()` function translates this [R]{.proglang} list into a
`pipeline.nix` file, which declaratively defines the entire workflow. The
package can then generate a visual representation of the pipeline's directed
acyclic graph (DAG), as shown in Figure 1.

@@

@@

### The Environment Definition (`gen-env.R`)

This script uses [rix]{.pkg} to define a `default.nix` file that contains all
necessary dependencies for [R]{.proglang}, [Julia]{.proglang}, and
[Python]{.proglang}.

```{=tex}
\begin{CodeInput}
R> # This script defines the polyglot environment our pipeline will run in.
R> library(rix)
R>
R> # Define the complete execution environment
R> rix(
+   # Pin the environment to a specific date for reproducibility.
+   date = "2025-10-14",
+
+   # R Packages
+   r_pkgs = c(
+     "ggplot2", "dplyr", "arrow", "rix", "rixpress", "quarto"
+   ),
+ 
+   # Julia Configuration
+   jl_conf = list(
+     jl_version = "lts",
+     jl_pkgs = c("Distributions", "DataFrames", "Arrow", "Random")
+   ),
+ 
+   # Python Configuration
+   py_conf = list(
+     py_version = "3.13",
+     py_pkgs = c("pandas", "scikit-learn", "xgboost", "pyarrow")
+   ),
+ 
+   project_path = ".",
+   overwrite = TRUE
+ )
\end{CodeInput}
```

### The RBC Model Simulation (`functions/functions.jl`)

This Julia script implements the state-space solution to the RBC model as a pure
function. The economic theory is based on @depaoli2009rbc.

@@

### The XGBoost Forecasting Model (`functions/functions.py`)

This Python script contains a series of modular functions to handle feature
engineering, data splitting, model training, and prediction.

@@

### The Visualization (`functions/functions.R`)

This R script uses [ggplot2]{.pkg} to create a plot comparing the model's output
against the XGBoost predictions.

@@

>>>>

The `rxp_populate()` function translates this [R]{.proglang} list into a
`pipeline.nix` file, which declaratively defines the entire workflow. The
package can then generate a visual representation of the pipeline's directed
acyclic graph:

@@

Figure @fig-dag can be generated by calling `rxp_ggdag()`.

@@

### The Environment Definition

This code snippet uses [rix]{.pkg} to generate a `default.nix` file that contains all
necessary dependencies for [R]{.proglang}, [Julia]{.proglang}, and
[Python]{.proglang}. It is typically saved in a file called `gen-env.R`:

```{=tex}
\begin{CodeInput}
# This script defines the polyglot environment our pipeline will run in.
library(rix)

# Define the complete execution environment
rix(
 # Pin the environment to a specific date for reproducibility.
 date = "2025-10-14",

 # R Packages
 r_pkgs = c(
   "ggplot2", "dplyr", "arrow", "rix", "rixpress", "quarto"
 ),

 # Julia Configuration
 jl_conf = list(
   jl_version = "lts",
   jl_pkgs = c("Distributions", "DataFrames", "Arrow", "Random")
 ),

 # Python Configuration
 py_conf = list(
   py_version = "3.13",
   py_pkgs = c("pandas", "scikit-learn", "xgboost", "pyarrow")
 ),

 project_path = ".",
 overwrite = TRUE
)
\end{CodeInput}
```

### The RBC Model Simulation

This Julia script implements the state-space solution to the RBC model as a pure
function, and is saved in a subfolder of the project named `functions` and in a script
`functions.jl`. The economic theory is based on @depaoli2009rbc.

@@

### The XGBoost Forecasting Model

This Python script contains a series of modular functions to handle feature
engineering, data splitting, model training, and prediction, and is saved in
`functions/functions.py`.

@@

### The Visualization

This R snippet uses [ggplot2]{.pkg} to create a plot comparing the model's output
against the XGBoost predictions and is saved in `functions/functions.R`.

@@

### Running the project

The structure of the project should look like this:

```bash
.
├── functions
│   ├── functions.jl
│   ├── functions.py
│   └── functions.R
├── gen-env.R
└── gen-pipeline.R

2 directories, 5 files
```

Assuming we are working on a system that has Nix installed, we can "drop" into
a temporary Nix shell with R and `{rix}` available:

```r
nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/main/inst/extdata/default.nix)"
```

Once the shell is ready, start R (by simply typing `R`) and run `source("gen-env.R")`
to generate the adequate `default.nix` file. This is the environment that we can use
to work interactively with the code while developing, and in which the pipeline will
be executed. We can then leave the R session and the shell entirely (by typing `exit`)
and type `nix-shell` to drop into the environment as defined by the `default.nix`.
The pipeline can now be built using `source("gen-pipeline.R")`.
